The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-07-20 script/de/script-de.tex
CT script, paper and pencil: header updated; defined sagecode float (de only) for sage code examples and added list at the end; improved the code of the first 5 Sage samples (en only yet)
CT script, paper and pencil: some small changes in de and en; translated chapter 2.5 from en to de; font of cover page adapted

% Formatierung der Fußnoten
\usepackage[flushmargin]{footmisc}
\usepackage{fullpage}
\usepackage{url}
% Literaturverzeichnis kapitelweise
\usepackage[labelstoglobalaux]{bibunits} % ohne labelstoglobalaux scheint man thebibilograpy und bibtex nicht mischen zu können

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{%\vskip -20 pt 
	{ Das CrypTool-Skript: \\ 
	Kryptographie, Mathematik und mehr} \\
        \vspace{30pt}
        {\large Hintergrundmaterial und Zusatzinformationen\\%*[-8pt]
         zum freien E-Learning-Programm CrypTool\\
         (10. Auflage -- zu CrypTool-Version 1.4.30)}
        \vspace{60pt}
}

\vspace{32pt}

\author{
(c) Prof.\ Bernhard Esslinger (Mitautor und Herausgeber), 1998-2009 \\
Frankfurt am Main\\*[1ex]
www.cryptool.org
}

@@

%\addcontentsline{toc}{section}{Index} %moved to style.ist
\printindex

>>>>

% Formatierung der Fußnoten
\usepackage[flushmargin]{footmisc}
\usepackage{fullpage}
\usepackage{url}
% Literaturverzeichnis kapitelweise
\usepackage[labelstoglobalaux]{bibunits} % ohne labelstoglobalaux scheint man thebibilograpy und bibtex nicht mischen zu können
\usepackage{float}
\floatstyle{ruled}
\newfloat{sagecode}{!h}{loc}[chapter]
\floatname{sagecode}{Sample}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{\vspace{25}
	{\Huge Das CrypTool-Skript: \\
	\vspace{5}
	Kryptographie, Mathematik und mehr} \\
        \vspace{20pt}
        {\LARGE Hintergrundmaterial und Zusatzinformationen\\
         zum freien E-Learning-Programm CrypTool} \\
	\vspace{15pt}
        {\large (10. Auflage -- veröffentlicht mit CrypTool-Version 1.4.30)}
        \vspace{180pt}
}

\author{
(c) Prof.\ Bernhard Esslinger (Mitautor und Herausgeber) \\
und das CrypTool Entwickler-Team, 1998-2009 \\
Frankfurt am Main\\*[1ex]
\url{www.cryptool.org}
}

@@

\clearpage
\listof{sagecode}{Verzeichnis der Sage-Programmbeispiele}
\addcontentsline{toc}{chapter}{Sage-Programmbeispiele}

%\addcontentsline{toc}{section}{Index} %moved to style.ist

\printindex

### 2009-07-20 script/en/chapters/paper_and_pencil.tex
CT script, paper and pencil: some small changes in de and en; translated chapter 2.5 from en to de; font of cover page adapted

The following chapter provides a broad overview of paper and pencil 
methods\footnote{Always added with links to further information.} each with
references to deeper information.
All techniques that people can apply manually to en- and decipher a message are 
embraced by this term. These methods were and still are especially popular with
secret services, as a writing pad and a pencil -- in contrast to electronic
aids -- are totally unsuspicious.

@@

\vskip +20 pt
%------------------------------------------------------------------------------
\section{Transposition ciphers}

@@

%------------------------------------------------------------------------------
% {\bf Column and row transposition}
\newpage % Damit die Tables immer direkt nach dem zugehörigen Item. 
\subsection[Column and row transposition ciphers]
    {Column and row transposition\footnotemark}
    \footnotetext{%
Most of the following methods can be simulated in CrypTool\index{CrypTool} 
under the menu {\bf Crypt/Decrypt \textbackslash{} Symmetric (classic) 
\textbackslash{} Permutation}.}

@@

%------------------------------------------------------------------------------
\newpage % Damit die Tables immer direkt nach dem zugehörigen Item. 
\section{Substitution ciphers}
\index{Substitution}

@@

%------------------------------------------------------------------------------
\section{Combining substitution and transposition}

@@

% ---------------------------------------------------------------------------
\newpage
\section{Examples using Sage}
\index{Sage}

@@

A first introduction to the CAS Sage can be found in the appendix \ref{s:appendix-using-sage}.

To make the sample code easier to understand, we used the structure and the naming conventions
shown in the graphics below:
\begin{itemize}
  \item Encryption consists of the two steps encoding and enciphering.
  \item Encoding adapts the letters in the given plaintext P to the case defined in the
        given alphabet, and all non-alphabet characters are filtered out.
  \item Enciphering creates the ciphertext C.
  \item Decryption also consists of two steps: deciphering and decoding.
  \item Decoding is only necessary if the symbols in the alphabet are no ASCII characters ???????????????? xxxxxxxxxxxxx.
\end{itemize}

@@

\newpage
\subsection{Transposition ciphers}

Tranposition ciphers are implemented in the class
\begin{center}
\verb!sage.crypto.classical.TranspositionCryptosystem!
\end{center}
To construct and work with a transposition cipher, we first need to
determine the alphabet that contains the symbols used to build the spave of our plaintext
and ciphertext. 
Typically, this alphabet will be the upper-case letters of the
English alphabet, which can be accessed via the function
\begin{center}
\verb!sage.monoids.string_monoid.AlphabeticStrings!
\end{center}
We then need to decide on the block length of a block permutation,
which is the length of the row vector to be used in encryption and
decryption. This row vector is our key, and it specifies a permutation
of a plaintext.

Here is a transposition cipher of block length 14, where every letter
in the plaintext is shifted to the left by two characters, with wrap
around. That is the encryption process. The decryption process is
shifting each letter of the ciphertext to the left by $14 - 2 = 12$
characters in order to recover the plaintext.

@@

\newpage
The next sample is also a simple column transposition. But now the code is a little bit more automated in generating the keys from the shift parameter.

@@

\newpage
As a more complicated example, we can use a permutation in the
encryption and decryption processes. If the block length of our
transposition cipher is $n$, then the encryption/decryption
key must be a permutation on $n$ symbols. The following
example uses the method \verb!random_key()! of the class
\verb!TranspositionCryptosystem!. So each call to
\verb!random_key()! produces a different key for encryption, which
affects the ciphertext. Note that your results may be different from
the following example.

@@

\newpage
The following example shows the key space of a transpositon cipher.

@@

Substitution cryptosystems are implemented in Sage in the class

\begin{center}
\verb!sage.crypto.classical.SubstitutionCryptosystem!
\end{center}

\noindent The following code sample uses Sage to construct a substitution
cipher with a random key. A random key can be generated using the
method \verb!random_key()! of the class
\texttt{SubstitutionCrypto\-system}. Different keys determine different
substitution ciphers. So each call to \verb!random_key()! returns
different results. First, we perform the encryption process with a
random key.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A   = AlphabeticStrings()
sage: S   = SubstitutionCryptosystem(A)
sage:
sage: P   = "Substitute this with something else better."
sage: key = S.random_key(); key
INZDHFUXJPATQOYLKSWGVECMRB
sage:
sage: # method encoding can be called from A or from T
sage: msg = A.encoding(P); msg
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: C   = S.enciphering(key, msg); C
WVNWGJGVGHGXJWCJGXWYQHGXJOUHTWHNHGGHS
sage:
sage: #.# We now decrypt the ciphertext to recover our plaintext.
sage:
sage: DC  = S.deciphering(key, C); DC
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: msg == DC
True
\end{Verbatim}
\caption{Monoalphabetic Substitution with randomly generated key}
\end{sagecode}

@@

The following example uses Sage to construct a Caesar
cipher. First, we demonstrate encryption using this cipher.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
....:          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: encrypt = S(key); encrypt
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: C       = encrypt(msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: #.# Next, we recover the plaintext.
sage:
sage: # decrypt message
sage: keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
....:             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: decrypt = S(keyInv); decrypt
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC      = decrypt(C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; key explicitely given, step-by-step approach)}
\end{sagecode}

\newpage
The next sample does the same, but the code is more sophisticated/automated/variable.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: keylen = len(A.gens()); keylen
26
sage: shift  = 3
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: S
Substitution cryptosystem on Free alphabetic string monoid on A-Z
sage: # key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
sage: #          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage: key = [(i+shift).mod(keylen) for i in range(keylen)];
sage: key = A(key); key
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: len(key)
26
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: C       = S.enciphering(key, msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: #.# Next, we recover the plaintext.
sage:
sage: # decrypt message
sage: # keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
sage: #             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) for i in range(keylen)];
sage: keyInv = A(keyInv); keyInv
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC     = S.enciphering(keyInv, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: # Just another way of decryption: Using "deciphering" with the key
sage: DC     = S.deciphering(key, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; keys generated)}
\end{sagecode}

@@

Here is a Sage example where the symbols are from the binary number
system. A monoalphabetic substitution cipher with a binary alphabet has very little
security: Because the plaintext/ciphertext alphabet has only the two elements
0 and 1, there are only two keys possible: (0 1) and (1 0).

@@

\newpage
However, our plaintext/ciphertext space  can be over a larger
alphabet, say the hexadecimal number system. Here is an example
using the hexadecimal number system as our substitution alphabet.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: A = HexadecimalStrings()
sage: S = SubstitutionCryptosystem(A)
sage: key = S.random_key(); key
2b56a4e701c98df3
sage: len(key)
16
sage: # Number of possible keys
sage: factorial(len(key))
20922789888000
sage: P   = "Working with a larger alphabet."
sage:
sage: msg = A.encoding(P); msg
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: C   = S.enciphering(key, msg); C
47e375e9e1efe75277e17ae052eb52e8eb75e7e47552ebe872e0ebe5e47a5f
sage: DC  = S.deciphering(key, C); DC
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: msg == DC
True
sage:
sage: #.# Geht das auch mit den Methoden des Alphabets oder der Chiffren ?????
sage: # Conversion hex back to ASCII:
sage: # - AlphabeticStrings() and HexadecimalStrings() don't have according methods.
sage: # - So we used Python:
sage: #   - repr(DC) converts the StringMonoidElement DC to a string
sage: #     still containing hex symbols;
sage: #   - binascii.a2b_hex converts this hex string to an ascii string.
sage: import binascii
sage: DDC = binascii.a2b_hex(repr(DC)); DDC
'Working with a larger alphabet.'
sage:
sage: P == DDC
True
\end{Verbatim}
\caption{Monoalphabetic substitution with a hexadecimal alphabet (and decoding in Python)}
\end{sagecode}

@@

The Vigen{\`e}re cipher is implemented in the Sage class

\begin{center}
\verb!sage.crypto.classical.VigenereCryptosystem!
\end{center}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # construct Vigenere cipher
sage: keylen = 14
sage: A = AlphabeticStrings()
sage: V = VigenereCryptosystem(A, keylen); V
Vigenere cryptosystem on Free alphabetic string monoid on A-Z of period 14
sage: #.# V.key_space() lieferte keinen Fehler, aber auch kein Ergebnis.
sage:
sage: # alternative: given key: key = A('ABCDEFGHIJKLMN'); key
sage: key = V.random_key(); key
WSSSEEGVVAARUD
sage: len(key)
14
sage:
sage: # encoding
sage: P = "The Vigenere cipher is polyalphabetic."
sage: len(P)
38
sage: msg = V.encoding(P); msg     # alternative: msg = A.encoding(P); msg
THEVIGENERECIPHERISPOLYALPHABETIC
sage:
sage: # encryption [2 alternative ways (in two steps or in one): both work]
sage: # encrypt = V(key); encrypt
sage: # C = encrypt(msg); C
sage: C   = V.enciphering(key, msg); C
PZWNMKKIZRETCSDWJAWTUGTALGBDXWLAG
sage:
sage: # decryption
sage: DC  = V.deciphering(key, C); DC
THEVIGENERECIPHERISPOLYALPHABETIC
sage: msg == DC
True
\end{Verbatim}
\caption{Vigen{\`e}re cipher}
\end{sagecode}

@@

The Hill~\cite{Hill1929,Hill1931} or matrix cipher is more
mathematically sophisticated than other ciphers mentioned in this
chapter. The encryption/decryption key of this cipher is an invertible
square matrix, and the plaintext/ciphertext is also a matrix. The
encryption and decryption processes use matrix operations. The Hill
cipher is implemented in the Sage class

\begin{center}
\verb!sage.crypto.classical.HillCryptosystem!
\end{center}

Here is an example where our plaintext/ciphertext space is the
capital letters of the English alphabet. In the Hill cipher, each
letter of this alphabet is assigned a unique integer modulo 26.

>>>>

The following chapter provides a broad overview of paper and pencil 
methods\footnote{%
The footnotes to this chapter describe how the methods can be
performd using CrypTool CrypTool 1.
Additionally the last sub chapter (\ref{PaP_Sage_samples})
contains example code using the computer algebra system Sage\index{Sage}.
                 }
each with references to deeper information.
All techniques that people can apply manually to en- and decipher a message are 
embraced by this term. These methods were and still are especially popular with
secret services, as a writing pad and a pencil -- in contrast to electronic
aids -- are totally unsuspicious.

@@

\newpage
%------------------------------------------------------------------------------
\section{Transposition ciphers}
\label{PaP_transposition_ciphers}
\index{Transposition}

@@

%------------------------------------------------------------------------------
% {\bf Column and row transposition}
\subsection[Column and row transposition ciphers]
    {Column and row transposition\footnotemark}
    \footnotetext{%
Most of the following methods can be simulated in CrypTool\index{CrypTool} 
under the menu {\bf Crypt/Decrypt \textbackslash{} Symmetric (classic) 
\textbackslash{} Permutation}.}

@@

%------------------------------------------------------------------------------
\section{Substitution ciphers}
\label{PaP_substitution_ciphers}
\index{Substitution}

@@

%------------------------------------------------------------------------------
\section{Combining substitution and transposition}
\label{PaP_combined_ciphers}
\index{Cascades}

@@

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\newpage
\section{Examples using Sage}
\label{PaP_Sage_samples}
\index{Sage!Code examples}
\index{Sage}

@@

\noindent A first introduction to the CAS Sage can be found in the appendix \ref{s:appendix-using-sage}.

\noindent To make the sample code\footnote{%
     Further examples with Sage can be found e.g. at:\\
     \href{http://www.sagemath.org/doc/constructions/linear_codes.html\#classical-ciphers}{\tt
           http://www.sagemath.org/doc/constructions/linear\_codes.html\#classical-ciphers}
                                          }
easier to understand, we used the structure and the naming conventions
shown in the graphics below:
\begin{itemize}
  \item Encryption consists of the two steps encoding and enciphering.
  \item Encoding adapts the letters in the given plaintext P to the case defined in the
        given alphabet, and all non-alphabet characters are filtered out.
  \item Enciphering creates the ciphertext C.
  \item Decryption also consists of two steps: deciphering and decoding.
  \item Decoding is only necessary if the symbols in the alphabet are no ASCII characters.
\end{itemize}

@@

% ---------------------------------------------------------------------------
\newpage
\subsection{Transposition ciphers}

Tranposition ciphers are implemented in the Sage class
\begin{center}
\verb!sage.crypto.classical.TranspositionCryptosystem!
\end{center}
To construct and work with a transposition cipher, we first need to
determine the alphabet that contains the symbols used to build the space of our plaintext
and ciphertext. 
Typically, this alphabet will be the upper-case letters of the
English alphabet, which can be accessed via the function
\begin{center}
\verb!sage.monoids.string_monoid.AlphabeticStrings!
\end{center}
We then need to decide on the block length of a block permutation,
which is the length of the row vector to be used in the simple columns
transposition. This row vector is our key, and it specifies a permutation
of a plaintext.

The following first example of transposition ciphers has block length 14,
and the key is build in a way, that every letter
in the plaintext is shifted to the right by two characters, with wrap
around at the end of the block. That is the encryption process. The decryption process is
shifting each letter of the ciphertext to the left by $14 - 2 = 12$.

@@

\newpage
The second example of transposition ciphers is also a simple shifting column transposition.
But now the code is a little bit more automated: The keys are generated from the shift parameter.

@@

\newpage
In the third example of transposition ciphers we use an arbitrary permutation as key in the
encryption and decryption processes in order to scramble the characters within each
block (block length = number of columns in a simple column transposition).
If the block length is $n$, then the key must be a permutation on $n$ symbols.
The following example uses the method \verb!random_key()! of the class
\verb!TranspositionCryptosystem!. Each call to \verb!random_key()! produces a
different key. Note that therefore your results (key and ciphertext) may be different
from the following example.

@@

\newpage
The fourth example of transposition ciphers additionally shows the key space of a simple
column transpositon.

@@

Substitution cryptosystems are implemented in Sage in the class
\begin{center}
\verb!sage.crypto.classical.SubstitutionCryptosystem!
\end{center}

\noindent The following code sample uses Sage to construct a substitution
cipher with a random key. A random key can be generated using the
method \verb!random_key()! of the class
\texttt{SubstitutionCrypto\-system}. Different keys determine different
substitution ciphers. So each call to \verb!random_key()! returns
different results.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A   = AlphabeticStrings()
sage: S   = SubstitutionCryptosystem(A)
sage:
sage: P   = "Substitute this with something else better."
sage: key = S.random_key(); key
INZDHFUXJPATQOYLKSWGVECMRB
sage:
sage: # method encoding can be called from A or from T
sage: msg = A.encoding(P); msg
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: C   = S.enciphering(key, msg); C
WVNWGJGVGHGXJWCJGXWYQHGXJOUHTWHNHGGHS
sage:
sage: #### We now decrypt the ciphertext to recover our plaintext.
sage:
sage: DC  = S.deciphering(key, C); DC
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: msg == DC
True
\end{Verbatim}
\caption{Monoalphabetic Substitution with randomly generated key}
\end{sagecode}

@@

The following example uses Sage to construct a Caesar
cipher.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
....:          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: encrypt = S(key); encrypt
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: C       = encrypt(msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: #### Next, we recover the plaintext.
sage:
sage: # decrypt message
sage: keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
....:             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: decrypt = S(keyInv); decrypt
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC      = decrypt(C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; key explicitely given, step-by-step approach)}
\end{sagecode}

\newpage
The second Caesar sample does the same, but the code is more sophisticated/automated/variable.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: keylen = len(A.gens()); keylen
26
sage: shift  = 3
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: S
Substitution cryptosystem on Free alphabetic string monoid on A-Z
sage: # key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
sage: #          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage: key = [(i+shift).mod(keylen) for i in range(keylen)];
sage: key = A(key); key
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: len(key)
26
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: C       = S.enciphering(key, msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: #### Next, we recover the plaintext.
sage:
sage: # decrypt message
sage: # keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
sage: #             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) for i in range(keylen)];
sage: keyInv = A(keyInv); keyInv
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC     = S.enciphering(keyInv, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: # Just another way of decryption: Using "deciphering" with the key
sage: DC     = S.deciphering(key, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; substitution keys are generated)}
\end{sagecode}

@@

In the following Sage example the symbols are from the binary number
system. A monoalphabetic substitution cipher with a binary alphabet has very little
security: Because the plaintext/ciphertext alphabet has only the two elements
0 and 1, there are only two keys possible: (0 1) and (1 0). Remark: In each key of a
substitution cipher all symbols of the alphabet have to appear once.

@@

\newpage
The second sample of a monoalphabetic substitution with symbols uses  a larger alphabet
as plaintext/ciphertext space as the first sample. Here the hexadecimal number system
is used as substitution alphabet.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: A = HexadecimalStrings()
sage: S = SubstitutionCryptosystem(A)
sage: key = S.random_key(); key
2b56a4e701c98df3
sage: len(key)
16
sage: # Number of possible keys
sage: factorial(len(key))
20922789888000
sage: P   = "Working with a larger alphabet."
sage:
sage: msg = A.encoding(P); msg
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: C   = S.enciphering(key, msg); C
47e375e9e1efe75277e17ae052eb52e8eb75e7e47552ebe872e0ebe5e47a5f
sage: DC  = S.deciphering(key, C); DC
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: msg == DC
True
sage:
sage: # Conversion hex back to ASCII:
sage: # - AlphabeticStrings() and HexadecimalStrings() don't have according methods.
sage: # - So we used Python:
sage: #   - repr(DC) converts the StringMonoidElement DC to a string
sage: #     still containing hex symbols;
sage: #   - binascii.a2b_hex converts this hex string to an ascii string.
sage: import binascii
sage: DDC = binascii.a2b_hex(repr(DC)); DDC
'Working with a larger alphabet.'
sage:
sage: P == DDC
True
\end{Verbatim}
\caption{Monoalphabetic substitution with a hexadecimal alphabet (and decoding in Python)}
\end{sagecode}

@@

The Vigen{\`e}re cipher is implemented in the Sage class
\begin{center}
\verb!sage.crypto.classical.VigenereCryptosystem!
\end{center}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # construct Vigenere cipher
sage: keylen = 14
sage: A = AlphabeticStrings()
sage: V = VigenereCryptosystem(A, keylen); V
Vigenere cryptosystem on Free alphabetic string monoid on A-Z of period 14
sage:
sage: # alternative: given key: key = A('ABCDEFGHIJKLMN'); key
sage: key = V.random_key(); key
WSSSEEGVVAARUD
sage: len(key)
14
sage:
sage: # encoding
sage: P = "The Vigenere cipher is polyalphabetic."
sage: len(P)
38
sage: msg = V.encoding(P); msg     # alternative: msg = A.encoding(P); msg
THEVIGENERECIPHERISPOLYALPHABETIC
sage:
sage: # encryption [2 alternative ways (in two steps or in one): both work]
sage: # encrypt = V(key); encrypt
sage: # C = encrypt(msg); C
sage: C   = V.enciphering(key, msg); C
PZWNMKKIZRETCSDWJAWTUGTALGBDXWLAG
sage:
sage: # decryption
sage: DC  = V.deciphering(key, C); DC
THEVIGENERECIPHERISPOLYALPHABETIC
sage: msg == DC
True
\end{Verbatim}
\caption{Vigen{\`e}re cipher}
\end{sagecode}

@@

The Hill~\cite{Hill1929,Hill1931} or matrix cipher is more
mathematically sophisticated than other ciphers mentioned in this
chapter. The encryption/decryption key of this cipher is an invertible
square matrix, and the plaintext/ciphertext is processed also as a matrix.
The encryption and decryption processes use matrix operations. The Hill
cipher is implemented in the Sage class
\begin{center}
\verb!sage.crypto.classical.HillCryptosystem!
\end{center}

In the following example our plaintext/ciphertext space is the
capital letters of the English alphabet. In the Hill cipher, each
letter of this alphabet is assigned a unique integer modulo 26.

### 2009-07-20 script/en/script-en.tex
paper and pencil: converted pseudo appendix in to section; defined sagecode float (en only) for sage code examples
sagecode float fine tuning
Put Sage code at the point where you want it. Use !h in LaTeX.
CT script, paper and pencil: some small changes in de and en; translated chapter 2.5 from en to de; font of cover page adapted

\usepackage[flushmargin]{footmisc}
\usepackage{fullpage}
\usepackage{url}
% Literaturverzeichnis kapitelweise
\usepackage[labelstoglobalaux]{bibunits} % ohne labelstoglobalaux scheint man thebibilograpy und bibtex nicht mischen zu können
%\bibliographyunit[\section] % bibunits für das ganze Dokument aktivieren
\defaultbibliography{references}
\defaultbibliographystyle{abbrv}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{ 
	The CrypTool Script: \\
	Cryptography, Mathematics and More \\
        \vspace{30pt}
        {\large	Background reading\\%*[-8pt] 
        for the free eLearning program CrypTool\\
       (10th edition -- distributed with CrypTool version 1.4.30)}
        \vspace{60pt}
}

\vspace{32pt}

\author{
Copyright (c) Prof.\ Bernhard Esslinger (co-author and editor) \\
and the CrypTool Development Team, 1998-2009 \\
Frankfurt am Main, Germany\\*[1ex]
\url{www.cryptool.org}
}

@@

>>>>

\usepackage[flushmargin]{footmisc}
\usepackage{fullpage}
\usepackage{url}
% Literaturverzeichnis kapitelweise
\usepackage[labelstoglobalaux]{bibunits} % ohne labelstoglobalaux scheint man thebibilograpy und bibtex nicht mischen zu können
%\bibliographyunit[\section] % bibunits für das ganze Dokument aktivieren
\defaultbibliography{references}
\defaultbibliographystyle{abbrv}
\usepackage{float}
\floatstyle{ruled}
\newfloat{sagecode}{!h}{loc}[chapter]
\floatname{sagecode}{Sample}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{ % Keine Leerzeilen, sonst Meldung: "Paragraph ended before \title was complete."
	{\Huge The CrypTool Script: \\
	 \vspace{5pt}
	Cryptography, Mathematics and More} \\
        \vspace{20pt}
	{\LARGE	Background reading\\ 
        for the free e-learning program CrypTool} \\
        \vspace{15pt}
        {\large (10th edition -- distributed with CrypTool version 1.4.30)}
        \vspace{180pt}
}

% Hier wird \vspace{32pt} einfach ignoriert
\author{
Copyright (c) Prof.\ Bernhard Esslinger (co-author and editor) \\
and the CrypTool Development Team, 1998-2009 \\
Frankfurt am Main, Germany\\*[1ex]
\url{www.cryptool.org}
}

@@

\clearpage
\listof{sagecode}{List of Sage Code Examples}
\addcontentsline{toc}{chapter}{List of Sage Code Examples}
